---
title: "Wang GBM–Ovarian T-cell harmonization"
subtitle: "Technical analysis report"
date: "`r format(Sys.time(), '%Y-%m-%d %H:%M %Z')`"
output:
  html_document:
    self_contained: true
    toc: true
    toc_float: true
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
library(data.table)
library(knitr)
run_dir <- Sys.getenv("CARTKFLS_RUN_DIR")
script_dir <- Sys.getenv("CARTKFLS_SCRIPT_DIR")
read_tab <- function(...) fread(file.path(run_dir, ...), na.strings = "NA")
embed_png <- function(path, alt_text) {
  if (!file.exists(path)) stop("Report image not found: ", path)
  uri <- knitr::image_uri(path)
  cat(sprintf(
    '<figure><img src="%s" alt="%s" style="max-width:100%%;height:auto;"><figcaption>%s</figcaption></figure>',
    uri, alt_text, alt_text
  ))
}
patient <- read_tab("sample_summaries", "harmonized_patient_summary.tsv")
counts <- read_tab("qc_tables", "harmonized_annotation_counts.tsv")
coverage <- rbindlist(list(
  read_tab("qc_tables", "GBM_signature_gene_coverage.tsv"),
  read_tab("qc_tables", "Ovarian_signature_gene_coverage.tsv")
))
validation <- read_tab("qc_tables", "harmonization_validation.tsv")
primary <- read_tab("association_results", "primary_total_t_associations.tsv")
cross <- read_tab("association_results", "cross_cancer_direction_comparison.tsv")
adjust <- read_tab("association_results", "one_feature_adjustment_models.tsv")
markers <- fread(file.path(script_dir, "marker_programs.tsv"))
```

# Answer first

This analysis places both datasets in a common T/NK framework while retaining the approved ovarian identities. T-cell subtype is represented as **CD8**, **conventional CD4**, **regulatory T**, or **unresolved T**; **NK remains a separate lineage**. Six non-exclusive biological programs—cytotoxicity, activation, naive/memory, exhaustion/dysfunction, interferon response, and stress—are retained as continuous UCell scores for every T/NK candidate cell.

The chromosome-fitness comparisons are exploratory cross-patient analyses. They test whether the association with total T-cell fraction changes after adding one composition or state feature at a time. They do not establish mediation or causality, and the small number of patients makes effect size and cross-dataset direction more informative than isolated nominal p-values.

# Data and annotation scope

```{r data-summary}
summary_table <- patient[, .(
  patients = uniqueN(patient_id),
  total_cells = sum(n_all_cells),
  T_cells = sum(n_t_cells),
  NK_cells = sum(n_nk_cells),
  median_T_fraction = median(total_t_fraction_all_cells, na.rm = TRUE)
), by = dataset]
kable(summary_table, digits = 4)
```

The ovarian subtype labels are mapped from the approved hierarchical annotations. GBM T/NK candidates are re-normalized and re-clustered within the T-cell compartment, with patient-aware Harmony integration when available. Cluster median lineage/subtype scores are then assigned against robust ovarian reference centroids. Clusters failing distance, lineage, margin, or minimum-size rules are labeled unresolved, with the reason retained per cell.

# Marker programs and scoring

```{r markers}
kable(markers[, .(signature, level, label, genes, citation, pmid, doi)], longtable = FALSE)
```

UCell rank scoring uses normalized RNA expression with `maxRank = 1500`. Raw UCell scores are the continuous analysis variables. Dataset-specific robust z-scores and primary-state labels are secondary browsing aids; they do not replace the continuous values.

```{r coverage}
kable(coverage[, .(dataset, signature, n_requested, n_observed, coverage_fraction)], digits = 3)
```

# Annotation and QC

```{r validation}
kable(validation)
```

```{r counts}
kable(counts[order(dataset, harmonized_lineage, harmonized_tcell_subtype, -N)])
```

```{r subtype-figure, results='asis'}
embed_png(file.path(run_dir, "figures", "subtype_composition_by_patient.png"), "Subtype composition by patient")
```

```{r program-figure, results='asis'}
embed_png(file.path(run_dir, "figures", "program_scores_by_patient.png"), "Continuous program scores by patient")
```

```{r gbm-umap-figure, results='asis'}
embed_png(file.path(run_dir, "figures", "GBM_tcell_subtype_umap.png"), "GBM T/NK reclustering and subtype assignments")
```

# Chromosome-fitness association analysis

For each selected patient ALFA-K landscape, frequent karyotypes are used to estimate three chromosome-level summaries: the marginal fitness slope per copy, a standardized linear effect, and a within-landscape Spearman correlation. Each summary is then correlated across patients with total T-cell fraction, T-cell subtype proportions, and mean continuous program scores. Ovarian split rows for `EOC153_omentum_1/2` are excluded from the primary patient-level analysis to avoid pseudo-replication; the combined `EOC153` row is retained.

```{r primary-results}
show_primary <- head(primary[order(p_value)], 30)
kable(show_primary, digits = 4)
```

```{r association-figure, results='asis'}
embed_png(file.path(run_dir, "figures", "chromosome_fitness_feature_associations.png"), "Chromosome-fitness associations with T-cell composition and state")
```

## Cross-cancer direction checks

```{r opposite}
if ("opposite_direction" %in% colnames(cross)) {
  show_cross <- head(cross[opposite_direction %in% TRUE], 40)
  if (nrow(show_cross)) kable(show_cross, digits = 3) else cat("No estimable cross-cancer associations had opposite signs.\n")
} else {
  cat("A complete two-dataset comparison was not estimable.\n")
}
```

## One-feature adjustment screens

Each model standardizes the chromosome effect, total T-cell fraction, and one added composition/state feature. Because the design has few patients and compositional predictors are dependent, features are added one at a time rather than combined into an overfit multivariable model.
Nominal p-values and Benjamini-Hochberg FDR values are both reported; the FDR family is defined within each cancer dataset and chromosome-effect metric across all tested chromosomes and added features.

```{r adjustments}
if (nrow(adjust)) {
  show_adjust <- head(adjust[order(-total_t_beta_attenuation_fraction, -delta_r_squared)], 40)
  kable(show_adjust, digits = 4)
}
```

# Interpretation limits

- GBM has a much smaller T-cell compartment than ovarian cancer; unresolved assignments are deliberately retained rather than forced.
- Program scores can co-occur. A primary state is only a convenience label and should not be treated as mutually exclusive biology.
- Cross-patient sample sizes are small, many chromosome-feature comparisons are tested, and patient-level observations are not evidence of cell-level causality.
- The landscape-derived chromosome summaries are operational marginal effects from the selected ALFA-K landscapes, not direct experimental chromosome perturbations.
- The one-feature adjustment screen identifies attenuation or strengthening consistent with composition/state dependence; it does not prove formal mediation.

# Reproducibility and files

The run directory contains harmonized Seurat objects, compressed cell metadata, patient/sample summaries, marker and gene-coverage documentation, QC tables, association tables, figures, execution logs, source-code checksums, input/container provenance, and completion sentinels. See `provenance/run_manifest.tsv` for an inventory.
